Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-07-14 ms/ms.Rnw
finalized ms for biorxiv

% Things to do before submission:
% Remove code to copy refs
% word count
% plain language summary

@@

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary:  190\\
Introduction:  \\
Materials and Methods:  \\
Results:  \\
Discussion:  \\
Acknowledgement:  \\
3 Figures and 1 Table, 1 Supplemental Figure

@@

  # Remove before publication
  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)

@@

\linenumbers
\setstretch{\stretchy}
% \pagenumbering{gobble}

@@

While the density and size of stomata have been researched extensively \citep[and references therin]{Sack_Buckley_2016}, the adaptive significance of stomatal distribution is less well understood. Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}. Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions. By creating a second parallel pathway for CO$_2$ diffusion within the mesophyll, amphistomy optimally supplies CO$_2$ \citep{Parkhurst_1978, Gutschick_1984b, Jones_1985}. Amphistomy is correlated with greater CO$_2$ diffusion \citep{Beerling_Kelly_1996} and higher photosynthetic rates \citep{Mckown_etal_2014}. These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}. On the other hand, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986, Buckley_etal_2015}. While transition to amphistomy is thus thought to increase transpiration, empirical studies suggest greater water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}. Hence, amphistomy appears to benefit a plant's carbon use relative to water loss and should be favored when CO$_2$ limits photosynthetic rate. The open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and groth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

To better understand the adapative significance of stomatal ratio, I asked three main questions:

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distibuted across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

@@

These observations from the British flora partially support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015} but also point out previously unknown subtleties. These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$). Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantitative variation. Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantiative indicator of vegetation type, rather than simply `open' or `closed'. A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

>>>>

@@

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 190\\
Introduction: 1179\\
Materials and Methods: 1176\\
Results: 433\\
Discussion: 1504\\
Acknowledgement: 24\\
3 Figures and 1 Table, 1 Supplemental Figure

@@

@@

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}

@@

While the density and size of stomata have been researched extensively \citep[and references therein]{Sack_Buckley_2016}, the adaptive significance of stomatal distribution is less well understood. Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}. Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions. By creating a second parallel pathway for CO$_2$ diffusion within the mesophyll, amphistomy optimally supplies CO$_2$ \citep{Parkhurst_1978, Gutschick_1984b, Jones_1985}. Amphistomy is correlated with greater CO$_2$ diffusion \citep{Beerling_Kelly_1996} and higher photosynthetic rates \citep{Mckown_etal_2014}. These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}. On the other hand, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986, Buckley_etal_2015}. While transition to amphistomy is thus thought to increase transpiration, empirical studies suggest greater water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}. Hence, amphistomy appears to benefit a plant's carbon use relative to water loss and should be favored when CO$_2$ limits photosynthetic rate. The open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, especially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shrubs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ families worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habitats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and growth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

To better understand the adaptive significance of stomatal ratio, I asked three main questions:

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SRmultReg}). Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SRmultReg}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary history. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

@@

These observations from the British flora partially support the hypothesis that amphistomy can be used a proxy for open habitat in paleoenvironment reconstruction \citep{Carpenter_1994, Jordan_etal_2014, Carpenter_etal_2015} but also point out previously unknown subtleties. These previous studies based their conclusions on data from Proteaceae, in which there is little quantitative variation in stomatal ratio; species are either completely hypostomatous ($\mathrm{SR_{propAd}} \approx 0$) or completely amphistomatous ($\mathrm{SR_{propAd}} \approx 0.5$). Stomatal ratio in British angiosperms is also bimodal \citep{Peat_Fitter_1994b}, but across many families there is also quantitative variation. Importantly, this means that quantitative variation in stomatal ratio may provide a more precise, quantitative indicator of vegetation type, rather than simply `open' or `closed'. A quantitative relationship between \el~and stomatal ratio has already been shown for herbaceous perennials \citep{Bucher_etal_2017}, but we now know that it holds among annuals (therophytes), subshrubs (chamaephytes), and, to a lesser extent, geophytes as well (Fig.~\ref{fig:SRmultReg}).

### 2017-07-14 ms/ms.tex
finalized ms for biorxiv

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary:  190\\
Introduction:  \\
Materials and Methods:  \\
Results:  \\
Discussion:  \\
Acknowledgement:  \\
3 Figures and 1 Table, 1 Supplemental Figure

\linenumbers
\setstretch{\stretchy}
% \pagenumbering{gobble}

>>>>

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary: 190\\
Introduction: 1179\\
Materials and Methods: 1176\\
Results: 433\\
Discussion: 1504\\
Acknowledgement: 24\\
3 Figures and 1 Table, 1 Supplemental Figure

\linenumbers
\setstretch{\stretchy}
%\pagenumbering{gobble}
